---
title: "Species Pull for Mill Springs Battlefield National Monument"
author:
- name: Matthew Van Scoyoc
  affiliation: |
    | NPS Species Inventory Program
    | NPS Inventory & Monitoring Division
    | 1201 Oakridge, Suite 150
    | Fort Collins, Colorado
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
params:
  UnitCode: MISP
  UnitName: Mill Springs Battlefield National Monument
---

```{r setup, echo=FALSE}
# Packages
if (isFALSE("here" %in% rownames(installed.packages()))) {
  install.packages("here", 
                   lib =  .libPaths()[1], 
                   repos = "https://cloud.r-project.org",
                   type = 'source', 
                   dependencies = TRUE, 
                   quiet = TRUE)
}
```

# Introduction

This [R Markdown](http://rmarkdown.rstudio.com) Notebook creates and documents a new sub-directory for a species pull request.
First, this document creates a sub-directory for the project. 
Then PARK_sppPull.Rmd is copied to the new sub-directory and renamed by replacing 'PARK' with the park unit code.
Appendices A and B are also copied to the new sub-directory. 
And finally, this script opens the .gitignore file so the new sub-directory can be added to the list of ignored sub-directories so the project is not synced to GitHub.

# New Project Sub-directory

This first chunk looks to see if there is already a sub-directory for the park in your project folder. 
If the park sub-directory does not exist, it creates the new sub-directory and copies and renames the Rmd files to the new sub-directory.
If the park sub-directory exists, you will get a warning message and the operation will not create the new sub-directory or copy the Rmd files.

```{r newDir, echo=TRUE, results='hide'}
# New sub-directory
parkDir <- paste(here::here(), params$UnitCode, sep = "/")

# Choose between PARK_sppPull.Rmd and PARK_sppPull_short.Rmd
myFile <- "PARK_sppPull.Rmd" 
myFile_short <- "PARK_sppPull_short.Rmd"

# Create new sub-directory and copy and rename files
if(!dir.exists(parkDir)){
  dir.create(params$UnitCode)
  file.copy("PARK_sppPull.Rmd", parkDir, overwrite = F, copy.date = T)
  file.rename(paste(parkDir, myFile, sep = "/"),
              paste(parkDir, paste0(params$UnitCode, "_sppPull.Rmd"), 
                    sep = "/"))
  file.copy("PARK_sppPull_short.Rmd", parkDir, overwrite = F, copy.date = T)
  file.rename(paste(parkDir, myFile_short, sep = "/"),
              paste(parkDir, paste0(params$UnitCode, "_sppPull_short.Rmd"), 
                    sep = "/"))
  file.copy("AppendixA Bird Phenology.Rmd", parkDir, 
            overwrite = F, copy.date = T)
  file.copy("AppendixB Species Heatmaps.Rmd", parkDir, 
            overwrite = F, copy.date = T)
} else(message("Directory already exists. Examine contents of directory before continuing to ensure data are not lost."))
```

# .gitignore

Now manually add the new park directory to the .gitignore file. 
There's a section at the bottom of the file that lists the park sub-directories.
Add "/PARK" to that list and substitute the park code for PARK, e.g., /MISP.

We do not want to sync the data or other products produced in a species data pull to GitHub due to sensitivity of the data and the large data files produced.

```{r}
file.edit(".gitignore")
```

# R Session Information
```{r session, results='hold'}
sessionInfo()
```
